Antigenic phenotypes and phylogenetic metrics improve forecasts of seasonal influenza A/H3N2 evolution

### 2020-08-04 manuscript/main.tex
Move supplemental figures and tables into main text
Promotes some supplemental figures to main figures and moves others to figure
supplements for the appropriate main figures. Also promotes most supplemental
tables to main tables.

We fit models for individual fitness metrics and combinations of metrics that we anticipated would be mutually beneficial.
For each model, we learned coefficient(s) that minimized the earth mover's distance between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Equation~\ref{equation_fitness_estimation}).
We evaluated model performance with time-series cross-validation such that better models reduced the earth mover's distance to the future on validation or test data (Supplemental Figures~\ref{sup_fig:cross_validation_for_simulated_populations} and \ref{sup_fig:cross_validation_for_natural_populations}).
The earth mover's distance to the future can never be zero, because each model makes predictions based on sequences available at the time of prediction and cannot account for new mutations that occur during the prediction interval.
We calculated the lower bound for each model's performance as the optimal distance to the future possible given the current sequences at each timepoint.
As an additional reference, we evaluated the performance of a ``naive'' model that predicted the future population would be identical to the current population.
We expected that the best models would consistently outperform the naive model and perform as close as possible to the lower bound.

@@

The long-term evolution of influenza H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \citep{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of H3N2-like viruses under positive and purifying selection for 40 years (Methods, Supplemental Figure~\ref{sup_fig:cross_validation_for_simulated_populations}).
These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural H3N2 HA (Supplemental Fig. \ref{sup_fig:simulated_h3n2_ha_phylogeny}, Supplemental Tables~\ref{sup_table:mutations_by_trunk_status_for_simulated_populations} and \ref{sup_table:mutations_by_trunk_status}).
We fit models to these simulated populations using all sequence-only fitness metrics.
As a positive control for our model framework, we also fit a model based on the true fitness of each strain as measured by the simulator.

@@

\begin{table}[htb]
  \begin{center}
    \scalebox{0.9}{
      \input{tables/simulated_model_selection.tex}
    }
    \caption{
      Simulated population model coefficients and performance on validation and test data ordered from best to worst by distance to the future in the validation analysis.
      Coefficients are the mean $\pm$ standard deviation for each metric in a given model across 33 training windows.
      Distance to the future (mean $\pm$ standard deviation) measures the distance in amino acids between estimated and observed future populations.
      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Figure~\ref{sup_fig:bootstrap_distributions_for_simulated_sample_3}).
      The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
      Test results are based on 18 timepoints not observed during model training and validation.
    }
    \label{table_simulated_model_selection}
  \end{center}
\end{table}

@@

\begin{figure}[htb]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.pdf}
  \caption{
    Simulated population model coefficients and distances to the future for individual biologically-informed fitness metrics and the best composite model.
    A) Coefficients and B) distances are shown per validation and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
  }
  \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}
  \end{center}
\end{figure}

Surprisingly, the composite model of epitope antigenic novelty and mutational load did not perform better than the individual mutational load model (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
The antigenic novelty fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \citep{Luksza:2014hj} that are not explicitly present in our simulations.
To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each strain's ancestor in the previous season.
This linear fitness metric slightly outperformed the antigenic novelty metric (Table~\ref{table_simulated_model_selection}).
Importantly, a composite model of the epitope ancestor and mutational load metrics outperformed all other epitope-based models and the individual mutational load model (Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated strains was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.

We hypothesized that a composite model of mutually beneficial metrics could better approximate the true fitness of simulated viruses than models based on individual metrics.
To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: mutational load and LBI.
This composite model outperformed both of its corresponding individual metric models with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperformed the naive model as often as the true fitness metric (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}, Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}).
The coefficients for mutational load and LBI remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
This small gain supports our hypothesis that multiple complementary metrics can produce more accurate models.

\begin{figure}
  \begin{center}
  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-normalized_fitness.pdf}
  \caption{
  Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
  A) The correlation of log estimated clade frequency fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$, and log observed clade frequency fold change, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$, shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future in the best model was in the top 20th percentile for 89\% of 18 timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  The corresponding results for the naive model are shown in Supplemental Figure~\ref{sup_fig:test_of_naive_model_for_simulated_populations}.
  }
  \label{fig:test_of_best_model_for_simulated_populations}
  \end{center}
\end{figure}

We validated the best performing model (true fitness) using two metrics that are relevant for practical influenza forecasting and vaccine design efforts.
First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by comparing observed fold change in clade frequencies, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$ and estimated fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$.
The model's estimated fold changes correlated well with observed fold changes (Pearson's $R^2 = 0.52$, Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}A).
The model also accurately predicted the growth of 87\% of growing clades and the decline of 58\% of declining clades.
Model forecasts were increasingly more accurate with increasing initial clade frequencies (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}C).
Next, we counted how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
The estimated best strain was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}B).
Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, Supplemental Figure~\ref{sup_fig:validation_of_best_model_for_simulated_populations}D).
In contrast, the naive model's forecasts of clade frequencies were considerably less accurate (Supplemental Figure~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}B).
However, the naive model's estimated closest strains to the future were consistently in the top fifth percentile of observed distances to the future and the correlation of its estimated percentile ranks and the observed ranks was strong (Spearman's $\rho^2 = 0.78$, Supplemental Figure~\ref{sup_fig:validation_of_naive_model_for_simulated_populations}C and D).
These results suggested that estimating a single closest strain to the future is a more tractable problem than estimating the future frequencies of clades.

Finally, we tested all of our models on out-of-sample data.
Specifically, we fixed the coefficients of each model to the average values across the validation period and applied the resulting models to the next 9 years of previously unobserved simulated data.
A standard expectation from machine learning is that models will perform worse on test data due to overfitting to training data.
Despite this expectation, we found that all models except for the individual epitope mutation models consistently outperformed the naive model across the out-of-sample data (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Supplemental Figure~\ref{sup_fig:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
The composite model of mutational load and LBI appeared to outperform the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively.
However, we did not find a significant difference between these models by bootstrap testing (Supplemental Table~\ref{sup_table:composite_vs_individual_model_comparison}) and could not rule out fluctuations in model performance across a relatively small number of data points.

As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
While observed and estimated clade frequency fold changes correlated more weakly for test data (Pearson's $R^2 = 0.14$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Figure~\ref{fig:test_of_best_model_for_simulated_populations}A).
We observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Supplemental Figure~\ref{fig:test_of_best_model_for_simulated_populations}C).
The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Figure~\ref{fig:test_of_best_model_for_simulated_populations}B).
Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}D).
The naive model performed comparatively well on these test data with all its estimated closest strains to the future in the top 20th percentile and a slightly higher correlation between observed and estimated percentile ranks than the true fitness model (Spearman's $\rho^2 = 0.82$, Supplemental Figure~\ref{sup_fig:test_of_naive_model_for_simulated_populations}).
These results confirmed that our approach of minimizing the distance between yearly populations could simultaneously capture clade-level dynamics of simulated influenza populations and identify individual strains that are most representative of future populations.
However, they also supported the earlier finding that clade frequency forecasts may be inherently more challenging than identification of the closest strain to the future.

@@

\begin{table}[htb]
  \begin{center}
    \scalebox{0.9}{
      \input{tables/natural_model_selection.tex}
    }
    \caption{
      Natural population model coefficients and performance on validation and test data ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Supplemental Figure~\ref{sup_fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
      Validation results are based on 23 timepoints.
      Test results are based on eight timepoints not observed during model training and validation.
    }
    \label{table_natural_model_selection}
  \end{center}
\end{table}

Next, we trained and validated models for individual fitness predictors using 25 years of natural H3N2 populations spanning from October 1, 1990 to October 1, 2015.
We held out strains collected after October 1, 2015 up through October 1, 2019 for model testing (Supplemental Figure~\ref{sup_fig:cross_validation_for_natural_populations}).
In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI antigenic novelty and DMS mutational effects.
We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
As antigenic drift is generally considered to be the primary evolutionary pressure on natural H3N2 populations \citep{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI antigenic novelty would be individually more predictive than mutational load or DMS mutational effects.
Previous research \citep{Neher:2014eu} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
As the earliest measurements from focus reduction assays (FRAs) date back to 2012, we could not train, validate, and test FRA antigenic novelty models in parallel with the HI antigenic novelty models.

\begin{figure}[htb]
  \begin{center}
  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-natural-populations.pdf}
  \caption{
    Natural population model coefficients and distances to the future for individual biologically-informed fitness metrics.
    A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
    The naive model's distance to the future (light gray) was 6.40 $\pm$ 1.36 AAs for validation timepoints and 6.82 $\pm$ 1.74 AAs for test timepoints.
    The corresponding lower bounds on the estimated distance to the future (dark gray) were 2.60 $\pm$ 0.89 AAs and 2.28 $\pm$ 0.61 AAs.
  }
  \label{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}
  \end{center}
\end{figure}

@@

Of the two metrics for antigenic drift, HI antigenic novelty consistently outperformed epitope antigenic novelty (Table~\ref{table_natural_model_selection}).
HI antigenic novelty estimated an average distance to the future of 6.01 $\pm$ 1.50 AAs and outperformed the naive model at 16 of 23 timepoints (70\%).
The coefficient for HI antigenic novelty remained stable across all timepoints (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
In contrast, epitope antigenic novelty estimated a distance of 7.13 $\pm$ 1.47 AAs and only outperformed the naive model at seven timepoints (30\%).
Epitope antigenic novelty was also the only metric whose coefficient started at a positive value (1.17 $\pm$ 0.03 on average prior to October 2009) and transitioned to a negative value through the validation period (-0.19 $\pm$ 0.34 on average for October 2009 and after).
This strong coefficient for the first half of training windows indicated that, unlike the results for simulated populations, the nonlinear antigenic novelty metric was historically an effective measure of antigenic drift.
The historical importance of the epitope sites used for this metric was further supported by the relative enrichment of mutations at these sites for the most successful ``trunk'' lineages of natural populations compared to side branch lineages (Supplemental Table~\ref{sup_table:mutations_by_trunk_status}).

>>>>

We fit models for individual fitness metrics and combinations of metrics that we anticipated would be mutually beneficial.
For each model, we learned coefficient(s) that minimized the earth mover's distance between HA amino acid sequences from the observed population one year in the future and the estimated population produced by the fitness model (Equation~\ref{equation_fitness_estimation}).
We evaluated model performance with time-series cross-validation such that better models reduced the earth mover's distance to the future on validation or test data (Figures~\ref{fig:cross_validation_for_simulated_populations} and \ref{fig:cross_validation_for_natural_populations}).
The earth mover's distance to the future can never be zero, because each model makes predictions based on sequences available at the time of prediction and cannot account for new mutations that occur during the prediction interval.
We calculated the lower bound for each model's performance as the optimal distance to the future possible given the current sequences at each timepoint.
As an additional reference, we evaluated the performance of a ``naive'' model that predicted the future population would be identical to the current population.
We expected that the best models would consistently outperform the naive model and perform as close as possible to the lower bound.

@@

The long-term evolution of influenza H3N2 hemagglutinin has been previously described as a balance between positive selection for substitutions that enable escape from adaptive immunity by modifying existing epitopes and purifying selection on domains that are required to maintain the protein's primary functions of binding and membrane fusion \citep{Bush:1999vj,Neher2013,Luksza:2014hj,Koelle:2015dh}.
To test the ability of our models to accurately detect these evolutionary patterns under controlled conditions, we simulated the long-term evolution of H3N2-like viruses under positive and purifying selection for 40 years (Methods, Figure~\ref{fig:cross_validation_for_simulated_populations}).
These selective constraints produced phylogenetic structures and accumulation of epitope and non-epitope mutations that were consistent with phylogenies of natural H3N2 HA (Figure \ref{fig:simulated_h3n2_ha_phylogeny}, Tables~\ref{table:mutations_by_trunk_status_for_simulated_populations} and \ref{table:mutations_by_trunk_status}).
We fit models to these simulated populations using all sequence-only fitness metrics.
As a positive control for our model framework, we also fit a model based on the true fitness of each strain as measured by the simulator.

@@

\begin{table}[htb]
  \begin{center}
    \scalebox{0.9}{
      \input{tables/simulated_model_selection.tex}
    }
    \caption{
      Simulated population model coefficients and performance on validation and test data ordered from best to worst by distance to the future in the validation analysis.
      Coefficients are the mean $\pm$ standard deviation for each metric in a given model across 33 training windows.
      Distance to the future (mean $\pm$ standard deviation) measures the distance in amino acids between estimated and observed future populations.
      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Figure~\ref{fig:bootstrap_distributions_for_simulated_sample_3}).
      The number of times (and percentage of total times) each model outperformed the naive model measures the benefit of each model over a model than estimates no change between current and future populations.
      Test results are based on 18 timepoints not observed during model training and validation.
    }
    \label{table_simulated_model_selection}
  \end{center}
\end{table}

@@

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-simulated-populations.pdf}
  \caption{
    Simulated population model coefficients and distances to the future for individual biologically-informed fitness metrics and the best composite model.
    A) Coefficients and B) distances are shown per validation and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
  }
  \label{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}

  \figsupp[Composite model coefficients and distances to the future for models fit to simulated populations.]{
    A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.}{\includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-simulated-populations.pdf}}\label{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}

\end{figure}

Surprisingly, the composite model of epitope antigenic novelty and mutational load did not perform better than the individual mutational load model (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}-Figure supplement~\ref{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
The antigenic novelty fitness metric assumes that antigenic drift is driven by nonlinear effects of previous host exposure \citep{Luksza:2014hj} that are not explicitly present in our simulations.
To understand whether positive selection at epitope sites might be better represented by a linear model, we fit an additional model based on an ``epitope ancestor'' metric that counted the number of epitope mutations since each strain's ancestor in the previous season.
This linear fitness metric slightly outperformed the antigenic novelty metric (Table~\ref{table_simulated_model_selection}).
Importantly, a composite model of the epitope ancestor and mutational load metrics outperformed all other epitope-based models and the individual mutational load model (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}-Figure supplement~\ref{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}).
From these results, we concluded that our method can accurately estimate the evolution of simulated populations, but that the fitness of simulated strains was dominated by purifying selection and only weakly affected by a linear effect of positive selection at epitope sites.

We hypothesized that a composite model of mutually beneficial metrics could better approximate the true fitness of simulated viruses than models based on individual metrics.
To this end, we fit an additional model including the best metrics from the mechanistic and clade growth categories: mutational load and LBI.
This composite model outperformed both of its corresponding individual metric models with an average distance to the future of 7.24 $\pm$ 1.66 AAs and outperformed the naive model as often as the true fitness metric (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}, Table~\ref{table:composite_vs_individual_model_comparison}).
The coefficients for mutational load and LBI remained relatively consistent across all validation timepoints, indicating that these fitness metrics were stable approximations of the simulator's underlying evolutionary processes.
This small gain supports our hypothesis that multiple complementary metrics can produce more accurate models.

\begin{figure}
  \includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-normalized_fitness.pdf}
  \caption{
  Test of best model for simulated populations (true fitness) using 9 years previously unobserved test data and fixed model coefficients.
  A) The correlation of log estimated clade frequency fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$, and log observed clade frequency fold change, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$, shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
  B) The rank of the estimated best strain based on its distance to the future in the best model was in the top 20th percentile for 89\% of 18 timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  The corresponding results for the naive model are shown in Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:test_of_naive_model_for_simulated_populations}.
  }
  \label{fig:test_of_best_model_for_simulated_populations}

  \figsupp[Validation of best model for simulated populations of H3N2-like viruses.]{
    A) The correlation of estimated and observed clade frequency fold changes shows the model's ability to capture clade-level dynamics without explicitly optimizing for clade frequency targets.
    B) The rank of the estimated best strain based on its distance to the future for 33 timepoints.
  The estimated best strain was in the top 20th percentile of observed closest strains for 100\% of timepoints, confirming that the model makes a good choice when forced to select a single representative strain for the future population.
  C) Absolute forecast error for clades shown in A by their initial frequency with a mean LOESS fit (solid black line) and 95\% confidence intervals (gray shading) based on 100 bootstraps.
  D) The correlation of all strains at all timepoints by the percentile rank of their observed and estimated distances to the future.
  The corresponding results for the naive model are shown in Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}.}{\includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3-normalized_fitness.pdf}}\label{figsupp:validation_of_best_model_for_simulated_populations}

  \figsupp[Validation of naive model for simulated populations of H3N2-like viruses.]{
    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3-naive.pdf}}\label{figsupp:validation_of_naive_model_for_simulated_populations}

  \figsupp[Test of naive model for simulated populations of H3N2-like viruses.]{
    Note that the naive model sets future frequencies to current frequencies such that there is no estimated fold change in frequencies for the first panel.}{\includegraphics[width=\textwidth]{figures/validation_figure_simulated-simulated_sample_3_test_tree-naive.pdf}}\label{figsupp:test_of_naive_model_for_simulated_populations}

\end{figure}

We validated the best performing model (true fitness) using two metrics that are relevant for practical influenza forecasting and vaccine design efforts.
First, we measured the ability of the true fitness model to accurately estimate dynamics of large clades (initial frequency $>15\%$) by comparing observed fold change in clade frequencies, $\log_{10}{\frac{x(t + \Delta{t})}{x(t)}}$ and estimated fold change, $\log_{10}{\frac{\hat{x}(t + \Delta{t})}{x(t)}}$.
The model's estimated fold changes correlated well with observed fold changes (Pearson's $R^2 = 0.52$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}A).
The model also accurately predicted the growth of 87\% of growing clades and the decline of 58\% of declining clades.
Model forecasts were increasingly more accurate with increasing initial clade frequencies (Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}C).
Next, we counted how often the estimated closest strain to the future population at any given timepoint ranked among the observed top closest strains to the future.
The estimated best strain was in the top first percentile of observed closest strains for half of the validation timepoints and in the top 20th percentile for 100\% of timepoints (Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}B).
Percentile ranks per strain based on their observed and estimated distances to the future correlated strongly across all strains and timepoints (Spearman's $\rho^2 = 0.87$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_best_model_for_simulated_populations}D).
In contrast, the naive model's forecasts of clade frequencies were considerably less accurate (Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}B).
However, the naive model's estimated closest strains to the future were consistently in the top fifth percentile of observed distances to the future and the correlation of its estimated percentile ranks and the observed ranks was strong (Spearman's $\rho^2 = 0.78$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:validation_of_naive_model_for_simulated_populations}C and D).
These results suggested that estimating a single closest strain to the future is a more tractable problem than estimating the future frequencies of clades.

Finally, we tested all of our models on out-of-sample data.
Specifically, we fixed the coefficients of each model to the average values across the validation period and applied the resulting models to the next 9 years of previously unobserved simulated data.
A standard expectation from machine learning is that models will perform worse on test data due to overfitting to training data.
Despite this expectation, we found that all models except for the individual epitope mutation models consistently outperformed the naive model across the out-of-sample data (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}, Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}, Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations}-Figure supplement~\ref{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_simulated_populations}, Table~\ref{table_simulated_model_selection}).
The composite model of mutational load and LBI appeared to outperform the true fitness metric with average distance to the future of 7.10 $\pm$ 1.19 compared to 7.38 $\pm$ 1.89, respectively.
However, we did not find a significant difference between these models by bootstrap testing (Table~\ref{table:composite_vs_individual_model_comparison}) and could not rule out fluctuations in model performance across a relatively small number of data points.

As with our validation dataset, we tested the true fitness model's ability to recapitulate clade dynamics and select optimal individual strains from the test data.
While observed and estimated clade frequency fold changes correlated more weakly for test data (Pearson's $R^2 = 0.14$), the accuracies of clade growth and decline predictions remained similar at 82\% and 53\%, respectively (Figure~\ref{fig:test_of_best_model_for_simulated_populations}A).
We observed higher absolute forecast errors in the test data with higher errors for clades between 40\% and 60\% initial frequencies (Figure~\ref{fig:test_of_best_model_for_simulated_populations}C).
The estimated best strain was higher than the top first percentile of observed closest strains for half of the test timepoints and in the top 20th percentile for 16 (89\%) of 18 of timepoints (Figure~\ref{fig:test_of_best_model_for_simulated_populations}B).
Observed and estimated strain ranks remained strongly correlated across all strains and timepoints (Spearman's $\rho^2 = 0.80$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}D).
The naive model performed comparatively well on these test data with all its estimated closest strains to the future in the top 20th percentile and a slightly higher correlation between observed and estimated percentile ranks than the true fitness model (Spearman's $\rho^2 = 0.82$, Figure~\ref{fig:test_of_best_model_for_simulated_populations}-Figure supplement~\ref{figsupp:test_of_naive_model_for_simulated_populations}).
These results confirmed that our approach of minimizing the distance between yearly populations could simultaneously capture clade-level dynamics of simulated influenza populations and identify individual strains that are most representative of future populations.
However, they also supported the earlier finding that clade frequency forecasts may be inherently more challenging than identification of the closest strain to the future.

@@

\begin{table}[htb]
  \begin{center}
    \scalebox{0.9}{
      \input{tables/natural_model_selection.tex}
    }
    \caption{
      Natural population model coefficients and performance on validation and test data ordered from best to worst by distance to the future in the validation analysis, as in Table~\ref{table_simulated_model_selection}.
      Distances annotated with asterisks (*) were significantly closer to the future than the naive model as measured by bootstrap tests (see Methods and Figure~\ref{fig:bootstrap_distributions_for_natural_sample_1_with_90_vpm_sliding}).
      Validation results are based on 23 timepoints.
      Test results are based on eight timepoints not observed during model training and validation.
    }
    \label{table_natural_model_selection}
  \end{center}
\end{table}

Next, we trained and validated models for individual fitness predictors using 25 years of natural H3N2 populations spanning from October 1, 1990 to October 1, 2015.
We held out strains collected after October 1, 2015 up through October 1, 2019 for model testing (Figure~\ref{fig:cross_validation_for_natural_populations}).
In addition to the sequence-only models we tested on simulated populations, we also fit models for our new fitness metrics based on experimental phenotypes including HI antigenic novelty and DMS mutational effects.
We hypothesized that both HI and DMS metrics would be assigned positive coefficients, as they estimate increased antigenic drift and beneficial mutations, respectively.
As antigenic drift is generally considered to be the primary evolutionary pressure on natural H3N2 populations \citep{Smith:2004jc,Bedford:2014bf,Luksza:2014hj}, we expected that epitope and HI antigenic novelty would be individually more predictive than mutational load or DMS mutational effects.
Previous research \citep{Neher:2014eu} and our simulation results also led us to expect that LBI and delta frequency would outperform other individual mechanistic metrics.
As the earliest measurements from focus reduction assays (FRAs) date back to 2012, we could not train, validate, and test FRA antigenic novelty models in parallel with the HI antigenic novelty models.

\begin{figure}[htb]
  \includegraphics[width=\textwidth]{figures/unadjusted-model-accuracy-and-coefficients-for-natural-populations.pdf}
  \caption{
    Natural population model coefficients and distances to the future for individual biologically-informed fitness metrics.
    A) Coefficients and B) distances are shown per validation timepoint (N=23) and test timepoint (N=8) as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
    The naive model's distance to the future (light gray) was 6.40 $\pm$ 1.36 AAs for validation timepoints and 6.82 $\pm$ 1.74 AAs for test timepoints.
    The corresponding lower bounds on the estimated distance to the future (dark gray) were 2.60 $\pm$ 0.89 AAs and 2.28 $\pm$ 0.61 AAs.
  }
  \label{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}

  \figsupp[Comparison of epitope-based models with knowledge of the future.]{
    Model coefficients and distances to the future for antigenic novelty models fit to natural populations.
    A) Coefficients and B) distances are shown per validation timepoint and test timepoint as in Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_simulated_populations_controls}.
    The epitope antigenic novelty model relies on previously published epitope sites \citep{Luksza:2014hj}.
    The ``oracle'' antigenic novelty model relies on sites of beneficial mutations that were manually identified from the entire training and validation time period (Methods).
    The improved performance of the ``oracle'' model indicates that the sequence-based antigenic novelty metric can be effective when sites of beneficial mutations are known prior to forecasting.}{\includegraphics[width=\textwidth]{figures/unadjusted-composite-model-accuracy-and-coefficients-for-natural-populations-epitope-vs-oracle.pdf}}\label{figsupp:unadjusted_composite_model_accuracy_and_coefficients_for_natural_populations_epitope_vs_oracle}

\end{figure}

@@

Of the two metrics for antigenic drift, HI antigenic novelty consistently outperformed epitope antigenic novelty (Table~\ref{table_natural_model_selection}).
HI antigenic novelty estimated an average distance to the future of 6.01 $\pm$ 1.50 AAs and outperformed the naive model at 16 of 23 timepoints (70\%).
The coefficient for HI antigenic novelty remained stable across all timepoints (Figure~\ref{fig:unadjusted_model_accuracy_and_coefficients_for_natural_populations}).
In contrast, epitope antigenic novelty estimated a distance of 7.13 $\pm$ 1.47 AAs and only outperformed the naive model at seven timepoints (30\%).
Epitope antigenic novelty was also the only metric whose coefficient started at a positive value (1.17 $\pm$ 0.03 on average prior to October 2009) and transitioned to a negative value through the validation period (-0.19 $\pm$ 0.34 on average for October 2009 and after).
This strong coefficient for the first half of training windows indicated that, unlike the results for simulated populations, the nonlinear antigenic novelty metric was historically an effective measure of antigenic drift.
The historical importance of the epitope sites used for this metric was further supported by the relative enrichment of mutations at these sites for the most successful ``trunk'' lineages of natural populations compared to side branch lineages (Table~\ref{table:mutations_by_trunk_status}).
